Unshuffling Permutations

### 2015-06-29 article.tex
Partie algebre (suite)

We say that a permutation $\pi$ {\em appears} in the shuffle
$\sigma \SHUFFLE \nu$ of two permutations $\sigma$ and $\nu$ if the
coefficient $\lambda_{\sigma, \nu}^\pi$ is different from zero.
Besides, a permutation $\pi$ is a {\em square} if there is a permutation
$\sigma$ such that $\pi$ appears in $\sigma \SHUFFLE \sigma$. In this
case, we say that $\sigma$ is a {\em square root} of $\pi$. Equivalently,
$\pi$ is a square if in the expansion of $\Delta(\pi)$, there is a
tensor $\sigma \otimes \sigma$ with a nonzero coefficient.
\medskip

@@

\begin{proposition} \label{prop:square_stability}
    Let $\pi$ be a square permutation and $\rho$ a square root of $\pi$.
    Then,
    \begin{enumerate}[label={\it (\roman*)},fullwidth]
        \item \label{item:square_stability_1}
        the permutation $\pi^\sim$ is a square and $\rho^\sim$
        is one of its square roots;
        \item \label{item:square_stability_2}
        the permutation $\bar \pi$ is a square and $\bar \rho$
        is one of its square roots;
        \item \label{item:square_stability_3}
        the permutation $\pi^{-1}$ is a square and $\rho^{-1}$
        is one of its square roots.
    \end{enumerate}
\end{proposition}
\begin{proof}
    First of all, since $\pi$ is a square permutation and $\rho$ is a
    square root of $\pi$, $\pi$ appears in $\rho \bullet \rho$. In
    an equivalent way, this amounts to say that $\rho \otimes \rho$ has
    a nonzero coefficient in the expansion of $\Delta(\pi)$. This
    implies that there are two disjoint sets of indexes $P_1$ and $P_2$
    such that $P_1 \uplus P_2 = [|\pi|]$ and
    \begin{equation} \label{equ:square_stability_same_std}
        \STD\left(\pi_{|P_1}\right) = \rho = \STD\left(\pi_{|P_2}\right).
    \end{equation}
    \smallskip

    To prove~\ref{item:square_stability_1}, set
    $P_1^\sim := \{|\pi| - i  + 1 : i \in P_1\}$ and
    $P_2^\sim := \{|\pi| - i  + 1 : i \in P_2\}$.
    Since~\eqref{equ:square_stability_same_std} holds, one has
    \begin{equation}
        \STD\left(\pi^\sim_{|P_1^\sim}\right) = \rho^\sim =
        \STD\left(\pi^\sim_{|P_2^\sim}\right).
    \end{equation}

>>>>

We say that a permutation $\pi$ {\em appears} in the shuffle
$\sigma \SHUFFLE \nu$ of two permutations $\sigma$ and $\nu$ if the
coefficient $\lambda_{\sigma, \nu}^\pi$ is different from zero.
Besides, a permutation $\pi$ is a {\em square} if there is a permutation
$\sigma$ such that $\pi$ appears in $\sigma \SHUFFLE \sigma$. In this
case, we say that $\sigma$ is a {\em square root} of $\pi$. Equivalently,
$\pi$ is a square with $\sigma$ as square root if and only if in the
expansion of $\Delta(\pi)$, there is a tensor $\sigma \otimes \sigma$
with a nonzero coefficient.
\medskip

@@

\begin{proposition} \label{prop:square_stability}
    Let $\pi$ be a square permutation and $\rho$ a square root of $\pi$.
    Then,
    \begin{enumerate}[label={\it (\roman*)},fullwidth]
        \item \label{item:square_stability_1}
        the permutation $\pi^\sim$ is a square and $\rho^\sim$
        is one of its square roots;
        \item \label{item:square_stability_2}
        the permutation $\bar \pi$ is a square and $\bar \rho$
        is one of its square roots;
        \item \label{item:square_stability_3}
        the permutation $\pi^{-1}$ is a square and $\rho^{-1}$
        is one of its square roots.
    \end{enumerate}
\end{proposition}
\begin{proof}
    First of all, since $\pi$ is a square permutation and $\rho$ is a
    square root of $\pi$, $\rho \otimes \rho$ has a nonzero coefficient
    in the expansion of $\Delta(\pi)$. This implies that there are two
    disjoint sets of indexes $P_1$ and $P_2$ such that
    $P_1 \uplus P_2 = [|\pi|]$ and, by denoting $\pi_{|P_1}$ by $u$ and
    $\pi_{|P_2}$ by $v$,
    \begin{equation} \label{equ:square_stability_same_std}
        \STD(u) = \rho = \STD(v).
    \end{equation}
    \smallskip

    To prove~\ref{item:square_stability_1}, set
    $P_1^\sim := \{|\pi| - i  + 1 : i \in P_1\}$,
    $P_2^\sim := \{|\pi| - i  + 1 : i \in P_2\}$,
    $u' := \pi^\sim_{|P_1^\sim}$, and $v' := \pi^\sim_{|P_2^\sim}$.
    Now, by construction, ${u'}^\sim = u$ and ${v'}^\sim = v$. Then,
    by~\eqref{equ:square_stability_same_std},
    $\STD(u') = \rho^\sim = \STD(v')$, so that $\rho^\sim$ is a square
    root of $\pi$.
    \smallskip

    Let us establish~\ref{item:square_stability_2}. Set
    $u' := \bar \pi_{|P_1}$ and $v' := \bar \pi_{|P_2}$. By definition
    of the operation $\bar\,$, for all indexes $i$ and $j$, we have
    $u_i < u_j$ if and only if $\bar u'_i > \bar u'_j$, the same
    property holding for $v$ and $v'$. Then,
    by~\eqref{equ:square_stability_same_std},
    $\STD(u') = \bar \rho = \STD(v')$, showing that $\bar \rho$ is a
    square root of $\bar \pi$.
    \smallskip

    Finally, let us show~\ref{item:square_stability_3}. Set
    $P^{-1}_1 := \{u_i : i \in [|u|]\}$,
    $P^{-1}_2 := \{v_i : i \in [|v|]\}$, $u' := \pi^{-1}_{|P^{-1}_1}$,
    and $v' := \pi^{-1}_{|P^{-1}_2}$.
